Bioinformatics analysis of whole-genome bisulfite-sequencing data

### 2015-05-06 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Edits from final proofread. How do I keep finding freaking typos!?

\begin{chapabstract}
I explain the bioinformatics analysis of whole-genome bisulfite-sequencing data, concentrating on the most widely used methylC-seq protocol. All data used in my thesis were generated using this protocol.

@@

The chapter concludes by introducing the \texttt{methtuple} software, a unique methylation caller for extracting methylation patterns at tuples of methylation loci. \texttt{methtuple} is critical for work in later chapters on co-methylation (Chapter \ref{chap:co-methylation}) but has wider application in facilitating downstream analyses of bisulfite-sequencing data.
\end{chapabstract}

@@

The `FastQC` software is a very useful tool for performing this first step ([http://www.bioinformatics.babraham.ac.uk/projects/fastqc/](http://www.bioinformatics.babraham.ac.uk/projects/fastqc/)). It produces summary graphs of many key measures such as base quality scores, read length distribution and sequence contamination. `FastQC` is a general purpose tool for performing quality control checks of high-throughput sequencing data. This means that some of its output must be interpreted with caution for bisulfite-sequencing data. For example, `FastQC` will report a warning (resp. error) if the relative frequency of the four bases differ by more than $10\%$ (resp. $20\%$). As noted in the `FastQC` documentation, such a warning/error should be ignored for bisulfite-sequencing data, owing to the inherent bias in their sequence composition.

@@

Read mapping is complicated by the bisulfite-treatment of the DNA. Following bisulfite-treatment, the DNA fragments are mostly composed of three bases rather than four, which means there are many more sequence mismatches between a read and its true mapping location. Simply using standard read mapping software and allowing for more mismatches would result in many reads mapping to multiple locations in the reference genome. Instead, a field of read mapping software dedicated to bisulfite-sequencing data has developed. Several review articles have summarised and compared the various approaches \citep[e.g.,][]{Chatterjee:2012ft, Krueger:2012ks, KundeRamamoorthy:2014fj}.

@@

There are biological and technical reasons why mapping against a reference genome can produce these errors. Biologically, if the sample contains sequences that are too genetically divergent from the reference genome then these sequences will be difficult, even impossible, to map. A particularly problematic class of sequences are those from repetitive regions of the genome. These repetitive sequences will map to multiple locations in the reference genome equally well. Furthermore, the number of times these repetitive sequences occur differs between the reference genome and sample's genome.

Technically, reads from Illumina sequencing are often too short to resolve the mapping location of these repetitive sequences. Resolving the mapping location of repetitive sequences can be achieved by using other sequencing technologies, such as Pacific Biosciences SMRT technology \citep{Flusberg:2010hra}, which produces longer reads.

Another source of technical error in read mapping is really due to sequencing error. A sequencing error can transform a uniquely mapping read to one that maps equally well to multiple locations or, worse still, a read that maps uniquely to a single, but incorrect, location. Sequencing errors can also corrupt a read so badly that it no longer can be mapped, which leads to that read being a false negative. In practice, most people try to mitigate these problems through their choice of parameters used by the read mapping software.

@@

PCR amplification of the input DNA is a common step in creating a library for high-throughput sequencing. PCR amplification is often required to ensure that there is a sufficient amount of DNA for the sequencer to properly work. Unfortunately, it can introduce biases into the library that result in some molecules being over-represented or under-represented compared to their true frequency. This means that when we sequence the library that we might sequence multiple fragments that are all copies of the same original piece of DNA, which gives a biased sampling of our sample's genome. These multiply-sequenced fragments are called _PCR duplicates_.

@@

In the datasets I have analysed, I have had to ignore up to 30 read-positions per read due to M-bias. Ideally, we would be able to remove the effects of M-bias by accounting for it in methylation calling rather than by simply excluding those read-positions entirely. One idea is to inversely weight methylation calls by the level of M-bias. A downside to this approach is that this would turn an otherwise binary methylation call into a continuous value between $0$ and $1$, with an attendant loss in interpretability. However, we could still compute the ubiquitous $\beta$-values, traditionally defined as the proportion of reads that are methylated at a locus, and use these in downstream inferences, without much loss in interpretation. The bigger problem with this approach is the increased computational complexity and cost, and this is why I have not further pursued this idea.

@@

Methylation calling is the process of calling each sequenced methylation locus as being either methylated or unmethylated[^no_call], as well as determining the _context_ or _type_ of each methylation event (i.e. CpG, CHG or CHH) based on the sequencing data and a reference DNA sequence. In principle, this is a simple process, however, this belies some complications, which I discuss in this section.

@@

1. Quantity: From a sequence containing $l$ methylation loci there are $l - \text{m} + 1$ $NIL = 0$ m-tuples. In contrast, there are $\binom{l}{\text{m}}$ $NIL \geq 0$ m-tuples. Obviously, $\binom{l}{\text{m}} \geq l - \text{m} + 1$, with strict inequality if $m \neq 1$ or $m \neq l$.
2. Interpretability: Results for m-tuples with $NIL = 0$ are simpler to interpret than when allowing $NIL \geq 0$. This is discussed in Chapter \ref{chap:co-methylation}.
3. Measurability: We cannot observe m-tuples where the methylation loci are far apart due to the read length limitations of the Illumina sequencing technology. This is true even when $NIL = 0$ but is more of an issue if we allow $NIL \geq 0$.

@@

`methtuple` performs methylation calling for a single `BAM` file generated by `Bismark`. The user is required to specify the size of the tuples (`--m`), and the methylation type (`--methylationType`) for each run of the program. There are many useful options to filter reads and read-positions. Apart from the standard quality filters, `methtuple` is careful when processing paired-end reads to only count the base from one of the reads in any overlapping paired-end reads to avoid double-counting these overlapping bases. `methtuple` also allows the user to filter out specific read-positions rather than wholesale filtering of the ends of reads. This is particularly useful for samples where there is a 'spike' in the M-bias plot, such as that shown in Figure \ref{fig:E18VA_M-bias}. Such a spike can be filtered out without also being forced to also filter out additional upstream read-positions that are not affected by M-bias.

@@

The running time of `methtuple` is proportional to the number of reads mapped to the chromosome, which is proportional to the length of the chromosome and its ploidy. The running time is largely independent of the tuple size (`-m`). The variation in running times within a chromosome is due to the number of reads generated per sample and the length of the reads, where the length of a paired-end read is defined as the sum of the mates' lengths. Samples with more reads take longer to process and samples sequenced with longer reads take longer because these contain more m-tuples.

>>>>

\begin{chapabstract}
This chapter explains the bioinformatics analysis of whole-genome bisulfite-sequencing data, concentrating on the most widely used methylC-seq protocol. All data used in my thesis were generated using this protocol.

@@

The chapter concludes by introducing the \texttt{methtuple} software that I wrote, a unique methylation caller for extracting methylation patterns at tuples of methylation loci. \texttt{methtuple} is critical for work in later chapters on co-methylation (Chapter \ref{chap:co-methylation}) but has wider application in facilitating downstream analyses of bisulfite-sequencing data.
\end{chapabstract}

@@

The `FastQC` software is a very useful tool for performing this first step ([http://www.bioinformatics.babraham.ac.uk/projects/fastqc/](http://www.bioinformatics.babraham.ac.uk/projects/fastqc/)). It produces summary graphs of many key measures such as base quality scores, read length distribution and sequence contamination. `FastQC` is a general purpose tool for performing quality control checks of high-throughput sequencing data. This means that some of its output must be interpreted with caution for bisulfite-sequencing data. For example, `FastQC` will report a warning (resp. error) if the relative frequency of the four bases differ by more than $10\%$ (resp. $20\%$). As noted in the `FastQC` documentation, such a warning/error should be ignored for bisulfite-sequencing data, owing to the inherent bias in its sequence composition.

@@

Read mapping is complicated by the bisulfite-treatment of the DNA. Following bisulfite-treatment, the DNA fragments are mostly composed of three bases rather than four, which means there are many more sequence mismatches between a read and its true mapping location. Simply using standard read mapping software and allowing for more mismatches would result in many reads mapping to multiple locations in the reference genome. Instead, a field of read mapping software dedicated to bisulfite-sequencing data has developed. Several review articles have summarised and compared the various approaches \citep{Chatterjee:2012ft, Krueger:2012ks, KundeRamamoorthy:2014fj}.

@@

There are biological and technical reasons why mapping against a reference genome can produce these errors. Biologically, if the sample contains sequences that are too genetically divergent from the reference genome then these sequences will be difficult, even impossible, to map. A particularly problematic class of sequences are those from repetitive regions of the genome. These repetitive sequences will map to multiple locations in the reference genome equally well. Furthermore, the number of times these repetitive sequences occur differs between the reference genome and the sample's genome.

Reads from Illumina sequencing are often too short to resolve the mapping location of these repetitive sequences. Resolving the mapping location of repetitive sequences can be achieved by using other sequencing technologies, such as Pacific Biosciences SMRT technology \citep{Flusberg:2010hra}, which produces longer reads.

Another source of technical error in read mapping is really due to sequencing error. A sequencing error can transform a uniquely mapping read to one that maps equally well to multiple locations or, worse still, a read that maps uniquely to a single, but incorrect, location. Sequencing errors can also corrupt a read so badly that it no longer can be mapped. In practice, most people try to mitigate these problems through their choice of parameters used by the read mapping software.

@@

PCR amplification of the input DNA is a common step in creating a library for high-throughput sequencing. PCR amplification is often required to ensure that there is a sufficient amount of DNA for the sequencer to properly work. Unfortunately, it can introduce biases into the library that results in some molecules being over-represented or under-represented compared to their true frequency. This means that when we sequence the library that we might sequence multiple fragments that are all copies of the same original piece of DNA, which gives a biased sampling of our sample's genome. These multiply-sequenced fragments are called _PCR duplicates_.

@@

In the datasets I have analysed, I have had to ignore up to 30 read-positions per read due to M-bias, a considerable loss of data. Ideally, we would be able to remove the effects of M-bias by accounting for it in methylation calling rather than by simply excluding those read-positions entirely. One idea is to inversely weight methylation calls by the level of M-bias. A downside to this approach is that this would turn an otherwise binary methylation call into a continuous value between $0$ and $1$, with an attendant loss in interpretability. However, we could still compute the ubiquitous $\beta$-values, traditionally defined as the proportion of reads that are methylated at a locus, and use these in downstream inferences, without much loss in interpretation. The bigger problem with this approach is the increased computational complexity and cost, and this is why I have not further pursued this idea.

@@

Methylation calling is the process of calling each sequenced methylation locus as being either methylated or unmethylated[^no_call], as well as determining the _context_ or _type_ of each methylation event (i.e. CpG, CHG or CHH) based on the sequencing data and a reference DNA sequence. In principle, this is a simple process, however, this belies some complications, which we discuss in this section.

@@

1. Quantity: From a sequence containing $l$ methylation loci there are $l - \text{m} + 1$ $NIL = 0$ m-tuples. In contrast, there are $\binom{l}{\text{m}}$ $NIL \geq 0$ m-tuples. Obviously, $\binom{l}{\text{m}} \geq l - \text{m} + 1$, with strict inequality if $m \neq 1$ or $m \neq l$.
2. Interpretability: Results for m-tuples with $NIL = 0$ are simpler to interpret than when allowing $NIL \geq 0$. This is discussed in Chapter \ref{chap:co-methylation}.
3. Measurability: We cannot observe methylation patterns from individual reads at m-tuples where the methylation loci are far apart due to the read length limitations of the Illumina sequencing technology. This is true even when $NIL = 0$ but is more of an issue if we allow $NIL \geq 0$.

@@

`methtuple` performs methylation calling for a single `BAM` file generated by `Bismark`. The user is required to specify the size of the tuples (`--m`), and the methylation type (`--methylationType`) for each run of the program. There are many useful options to filter reads and read-positions. Apart from the standard quality filters, `methtuple` is careful when processing paired-end reads to only count the base from one of the reads in any overlapping paired-end reads to avoid double-counting the bases in the overlapping region. `methtuple` also allows the user to filter out specific read-positions rather than wholesale filtering of the ends of reads. This is particularly useful for samples where there is a 'spike' in the M-bias plot, such as that shown in Figure \ref{fig:E18VA_M-bias}. Such a spike can be filtered out without also being forced to also filter out additional upstream read-positions that are not affected by M-bias.

@@

The running time of `methtuple` is proportional to the number of reads mapped to the chromosome, which is proportional to the length of the chromosome and its copy number. The running time is largely independent of the tuple size (`-m`). The variation in running times within a chromosome is due to the number of reads generated per sample and the length of the reads, where the length of a paired-end read is defined as the sum of the mates' lengths. Samples with more reads take longer to process and samples sequenced with longer reads take longer because these contain more m-tuples.
